---
title: "Advanced stablr Workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced stablr Workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4
)
set.seed(42)
```

## Overview

The introductory and multi-omic vignettes focus on the core Gaussian and
binomial workflows. This vignette covers additional `stablr` capabilities that
are easy to miss when reading only the quick-start paths:

| Topic | Functions |
|-------|-----------|
| Survival STABL | `stabl_fit(family = "cox")` |
| Multinomial classification | `stabl_fit(family = "multinomial")` |
| Artificial features | `artificial_type = "random_permutation"`, `"knockoff"`, `"knockoff_equi"`, `"knockoff_mvr"`, or `NULL` |
| Grouped bootstrap | `groups`, `group_bootstrap_indices()` |
| Reproducibility metrics | `jaccard_similarity()`, `adjusted_similarity()` |
| Disk export | `export_stabl_to_csv()`, `save_stabl_results()` |
| Outer multi-omic CV | `stabl_multiomic_cv()` |
| Cooperative diagnostics | `get_cooperative_features()`, `plot()` on native fits |

Examples use small simulated data so the vignette remains fast during
`R CMD check`.

```{r load}
library(stablr)
```

---

## 1. Cox proportional-hazards STABL

For time-to-event outcomes, pass a `survival::Surv` object as `y` and set
`family = "cox"`. Row names of `y` must match `rownames(x)`.

```{r cox-setup}
library(survival)

n <- 80
p <- 15
ids <- paste0("s", seq_len(n))
x_cox <- matrix(
  rnorm(n * p), nrow = n,
  dimnames = list(ids, paste0("g", seq_len(p)))
)
# Plant signal in first two features
linpred <- 0.9 * x_cox[, 1] - 0.7 * x_cox[, 2]
y_cox <- Surv(time = rexp(n, exp(linpred)), event = rbinom(n, 1, 0.75))
rownames(y_cox) <- ids

lambda_cox <- auto_lambda_grid(x_cox, y_cox, family = "cox", n_lambda = 8)
```

```{r cox-fit}
fit_cox <- stabl_fit(
  x               = x_cox,
  y               = y_cox,
  lambda_grid     = lambda_cox,
  family          = "cox",
  n_bootstraps    = 40L,
  artificial_type = "random_permutation",
  random_state    = 42L
)

get_feature_names_out(fit_cox)
head(sort(get_importances(fit_cox), decreasing = TRUE), 6)
```

---

## 2. Multinomial STABL

Multinomial outcomes use a named factor `y`. The stability matrix has one row
per feature; importances aggregate across classes.

```{r multinomial}
n <- 90
p <- 12
ids <- paste0("s", seq_len(n))
x_multi <- matrix(
  rnorm(n * p), nrow = n,
  dimnames = list(ids, paste0("m", seq_len(p)))
)
y_multi <- factor(sample(c("A", "B", "C"), n, replace = TRUE), levels = c("A", "B", "C"))
names(y_multi) <- ids
# Weak class signal from first feature
x_multi[y_multi == "B", 1] <- x_multi[y_multi == "B", 1] + 1.5
x_multi[y_multi == "C", 1] <- x_multi[y_multi == "C", 1] - 1.5

lambda_multi <- auto_lambda_grid(x_multi, y_multi, family = "multinomial", n_lambda = 6)

fit_multi <- stabl_fit(
  x               = x_multi,
  y               = y_multi,
  lambda_grid     = lambda_multi,
  family          = "multinomial",
  n_bootstraps    = 30L,
  artificial_type = "random_permutation",
  random_state    = 1L
)

get_feature_names_out(fit_multi)
```

---

## 3. Knockoff artificial features

Supported artificial-feature settings are `"random_permutation"`, `"knockoff"`,
`"knockoff_equi"`, `"knockoff_mvr"`, and `NULL`.  Use
`artificial_type = "knockoff"` for fixed-X knockoffs.  Use
`artificial_type = "knockoff_equi"` or `"knockoff_mvr"` for Model-X Gaussian
knockoffs when the optional `knockoff` package is installed. All three
knockoff modes support at most 3,000 predictors and fail closed above that
limit; no chunked construction is claimed to be scientifically valid.

```{r knockoff-check, include = FALSE}
has_knockoff <- requireNamespace("knockoff", quietly = TRUE)
if (!has_knockoff) {
  message("Install the knockoff package to evaluate the knockoff example.")
}
```

```{r knockoff-fit, eval = has_knockoff}
n <- 60
p <- 20
ids <- paste0("s", seq_len(n))
x_ko <- matrix(
  rnorm(n * p), nrow = n,
  dimnames = list(ids, paste0("k", seq_len(p)))
)
y_ko <- setNames(as.integer(x_ko[, 1] + rnorm(n) > 0), ids)

lambda_ko <- auto_lambda_grid(x_ko, y_ko, family = "binomial", n_lambda = 8)

fit_ko <- stabl_fit(
  x               = x_ko,
  y               = y_ko,
  lambda_grid     = lambda_ko,
  family          = "binomial",
  n_bootstraps    = 30L,
  artificial_type = "knockoff_equi",
  random_state    = 7L
)

get_feature_names_out(fit_ko)
```


---

## 4. Grouped bootstrap (repeated measures)

When several rows belong to the same subject or batch, pass a `groups` vector so
bootstrap resampling keeps whole groups together. This prevents leakage across
related samples.

```{r grouped-bootstrap}
n <- 48
p <- 10
ids <- paste0("s", seq_len(n))
groups <- setNames(rep(paste0("subj", seq_len(12)), each = 4), ids)
x_grp <- matrix(
  rnorm(n * p), nrow = n,
  dimnames = list(ids, paste0("f", seq_len(p)))
)
y_grp <- setNames(rnorm(n) + 0.5 * x_grp[, 1], ids)

# Inspect one grouped draw (whole subjects only)
idx <- group_bootstrap_indices(y_grp, groups = groups, n_subsamples = 24L, seed = 1L)
unique(groups[idx])

lambda_grp <- auto_lambda_grid(x_grp, y_grp, family = "gaussian", n_lambda = 6)

fit_grp <- stabl_fit(
  x               = x_grp,
  y               = y_grp,
  groups          = groups,
  lambda_grid     = lambda_grp,
  family          = "gaussian",
  n_bootstraps    = 25L,
  artificial_type = "random_permutation",
  random_state    = 3L
)

length(get_feature_names_out(fit_grp))
```

The same `groups` argument is forwarded by `stabl_multiomic_train_validate()`
and `stabl_multiomic_cv()` for outer fold construction.

---

## 5. Reproducibility across repeated fits

Selection-stability metrics compare feature lists from repeated STABL runs.
Jaccard similarity is the simplest overlap measure; adjusted similarity
corrects for chance overlap when set sizes differ.

```{r reproducibility}
run_once <- function(seed) {
  stabl_fit(
    x = x_grp, y = y_grp, groups = groups,
    lambda_grid = lambda_grp,
    family = "gaussian",
    n_bootstraps = 20L,
    artificial_type = "random_permutation",
    random_state = seed
  )
}

fit_a <- run_once(11L)
fit_b <- run_once(22L)
sel_a <- get_feature_names_out(fit_a)
sel_b <- get_feature_names_out(fit_b)

jaccard_similarity(sel_a, sel_b)
adjusted_similarity(sel_a, sel_b, nb_total_elements = ncol(x_grp))

# Pairwise matrix across several seeds
fits <- lapply(11:14, run_once)
sels <- lapply(fits, get_feature_names_out)
jaccard_matrix(sels)
```

---

## 6. Export results to disk

`export_stabl_to_csv()` writes stability-score tables for a single-omic fit.
`save_stabl_results()` bundles scores, diagnostics, selected-feature tables, and feature-distribution figures for a full analysis folder. It writes `STABL scores.csv`, `Max STABL scores.csv`, `FDR Graph.<fmt>`, `Stability Path.<fmt>`, `Selected Features/Selected features.csv`, and `Selected Features/Feature distributions.<fmt>` when the relevant diagnostics and selected features are available.

```{r export, eval = FALSE}
tmp <- tempfile("stablr_export")
dir.create(tmp)

export_stabl_to_csv(fit_grp, path = file.path(tmp, "grp"))
list.files(file.path(tmp, "grp"))

save_stabl_results(
  object     = fit_grp,
  path       = file.path(tmp, "bundle"),
  x          = x_grp,
  y          = y_grp,
  task_type  = "regression",
  figure_fmt = "png",
  override   = TRUE
)
```

---

## 7. Outer cross-validation for multi-omic workflows

When no fixed validation split exists, `stabl_multiomic_cv()` runs STABL (and
optional fusion branches) on rotating outer folds. Cooperative fusion is
supported here; nested CV (`stabl_multiomic_nested_cv()`) does not forward
cooperative mode.

```{r outer-cv}
n <- 50
ids <- paste0("s", seq_len(n))
x_a <- matrix(rnorm(n * 8), nrow = n, dimnames = list(ids, paste0("a", 1:8)))
x_b <- matrix(rnorm(n * 8), nrow = n, dimnames = list(ids, paste0("b", 1:8)))
y_cv <- setNames(0.6 * x_a[, 1] - 0.4 * x_b[, 2] + rnorm(n, sd = 0.3), ids)

lambda_cv <- list(
  omic_a = auto_lambda_grid(x_a, y_cv, family = "gaussian", n_lambda = 5),
  omic_b = auto_lambda_grid(x_b, y_cv, family = "gaussian", n_lambda = 5)
)

cv_fit <- stabl_multiomic_cv(
  x_list          = list(omic_a = x_a, omic_b = x_b),
  y               = y_cv,
  lambda_grid     = lambda_cv,
  v               = 3L,
  family          = "gaussian",
  n_bootstraps    = 15L,
  artificial_type = "random_permutation",
  random_state    = 5L,
  cooperative_fusion = TRUE,
  rho             = c(0, 0.3),
  cooperation_nfolds = 3L
)

head(cv_fit$diagnostics)
if (!is.null(cv_fit$fold_results[[1]]$cooperative_fusion)) {
  get_cooperative_features(cv_fit$fold_results[[1]])
}
```

---

## 8. Cooperative coefficient paths (native engine)

The built-in cooperative engine registers `plot.multiview()` for fitted
cooperative models. After a cooperative workflow, the selected fit object lives
in `$cooperative_fusion$fit`.

```{r cooperative-plot, fig.alt = "Cooperative coefficient paths by view"}
ool <- load_ool_data(split = "train")
lambda_ool <- list(
  cytof = auto_lambda_grid(ool$x_list$cytof, ool$y, n_lambda = 6),
  proteomics = auto_lambda_grid(ool$x_list$proteomics, ool$y, n_lambda = 6)
)

coop_fit <- stabl_multiomic_train_validate(
  x_train_list       = ool$x_list,
  y_train            = ool$y,
  lambda_grid        = lambda_ool,
  family             = "gaussian",
  n_bootstraps       = 12L,
  artificial_type    = "random_permutation",
  random_state       = 42L,
  cooperative_fusion = TRUE,
  rho                = 0.3,
  cooperation_nfolds = 3L
)

cf <- coop_fit$cooperative_fusion
get_cooperative_diagnostics(coop_fit)
plot(cf$fit)
```

---

## Related vignettes

- `vignette("stablr-intro")` — single-omic STABL, learners, diagnostic plots.
- `vignette("stablr-multiomic")` — early and late fusion on OOL data.
- `vignette("stablr-cooperative")` — cooperation strength `rho`, validation vs
  CV tuning, binomial cooperative fusion.
- `docs/PYTHON_TO_R_MAPPING.md` — mapping to the Python tutorial API.
