---
title: "Research protocol: TCGA multi-omic biomarker discovery"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Research protocol: TCGA multi-omic biomarker discovery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 6,
  fig.height = 4,
  cache      = FALSE,
  cache.path = "stablr-tcga-cache/",
  # eval=FALSE: vignette requires large TCGA data and long-running fits;
  # gate from R CMD check and CI. Run interactively or via pkgdown::build_site().
  eval       = FALSE
)
```

> **Non-executable research protocol.** No TCGA result in this page is package
> or release evidence. Code chunks are set to `eval = FALSE` because the data
> are external and the fits are long-running. Execute the protocol only after
> pinning the dataset, candidate tarball, settings, and output manifest.

## Overview

Breast cancer subtype is a molecular label, but it is also a test of
integration.  The same tumour is measured by mRNA and miRNA assays.  Some
signals should be visible in one view, some in the other, and some only when the
views are allowed to compete or cooperate.

This vignette translates the **mixOmics N-Integration Chapter 6** case study
into the `stablr` idiom.  The original mixOmics workflow uses sparse
multi-block Discriminant Analysis (sgccaDA) to classify breast cancer subtypes.
Here we ask a narrower question: which mRNA and miRNA features robustly
distinguish Basal-like tumours from all other subtypes?

The proposed analysis uses the `breast.TCGA` train/test split from `mixOmics`,
with compact settings shown only to specify the workflow. No result is reported
here. A publication analysis requires a separately predeclared and hashed run.

**Dataset:** `breast.TCGA` from the `mixOmics` package.

- Training: mRNA and miRNA measurements on the same tumour samples.
- Test: held-out mRNA and miRNA measurements with matching feature sets.
- Outcome: PAM50 subtype binarised to **Basal vs. non-Basal**.
- Protein data is excluded to match the two-omic scope used here.

## Prerequisites

```{r prerequisites}
if (!requireNamespace("mixOmics", quietly = TRUE)) {
  stop(
    "This vignette requires mixOmics (Bioconductor). ",
    "Install with: BiocManager::install('mixOmics')"
  )
}
library(stablr)
library(mixOmics)
```

## 1. Load and prepare data

The train and test matrices are already aligned within the `mixOmics` object.
We extract mRNA and miRNA, then turn the PAM50 subtype into a binary Basal vs
non-Basal outcome.

```{r load-data}
data(breast.TCGA)

X_train_mrna  <- breast.TCGA$data.train$mrna
X_train_mirna <- breast.TCGA$data.train$mirna
X_test_mrna   <- breast.TCGA$data.test$mrna
X_test_mirna  <- breast.TCGA$data.test$mirna

cat("Training mRNA:  ", nrow(X_train_mrna),  "x", ncol(X_train_mrna),  "\n")
cat("Training miRNA: ", nrow(X_train_mirna), "x", ncol(X_train_mirna), "\n")
cat("Test mRNA:      ", nrow(X_test_mrna),   "x", ncol(X_test_mrna),   "\n")
cat("Test miRNA:     ", nrow(X_test_mirna),  "x", ncol(X_test_mirna),  "\n")
```

```{r outcome}
subtype_train <- breast.TCGA$data.train$subtype
subtype_test  <- breast.TCGA$data.test$subtype

y_train <- as.integer(subtype_train == "Basal")
y_test  <- as.integer(subtype_test  == "Basal")

names(y_train) <- rownames(X_train_mrna)
names(y_test)  <- rownames(X_test_mrna)

cat("Training: ", sum(y_train), "Basal /", sum(y_train == 0), "non-Basal\n")
cat("Test:     ", sum(y_test),  "Basal /", sum(y_test == 0),  "non-Basal\n")
```

## 2. Per-omic STABL

The first pass treats each assay as its own experiment.  This tells us whether
mRNA and miRNA each contain stable Basal signal before the integrated model is
allowed to mix them.  The vignette uses 30 bootstraps and 12 lambda values to
keep runtime bounded; use 500-1000 bootstraps for final reporting.

```{r mrna-fit}
lambda_mrna <- auto_lambda_grid(
  X_train_mrna, y_train,
  family   = "binomial",
  n_lambda = 12
)

fit_mrna <- stabl_fit(
  x               = X_train_mrna,
  y               = y_train,
  lambda_grid     = lambda_mrna,
  family          = "binomial",
  n_bootstraps    = 30L,
  artificial_type = "random_permutation",
  random_state    = 42L
)
fit_mrna
```

```{r mirna-fit}
lambda_mirna <- auto_lambda_grid(
  X_train_mirna, y_train,
  family   = "binomial",
  n_lambda = 12
)

fit_mirna <- stabl_fit(
  x               = X_train_mirna,
  y               = y_train,
  lambda_grid     = lambda_mirna,
  family          = "binomial",
  n_bootstraps    = 30L,
  artificial_type = "random_permutation",
  random_state    = 42L
)
fit_mirna
```

```{r selected-per-omic}
cat("mRNA selected feature names:\n")
print(get_feature_names_out(fit_mrna))

cat("\nmiRNA selected feature names:\n")
print(get_feature_names_out(fit_mirna))
```

```{r stability-paths, fig.alt = "mRNA and miRNA stability paths"}
print(plot_stabl_path(fit_mrna,  title = "mRNA - stability path"))
print(plot_stabl_path(fit_mirna, title = "miRNA - stability path"))
```

## 3. Integrated multi-omic pipeline

The integrated workflow repeats the per-omic fits inside a common pipeline,
adds early fusion on the concatenated matrix, and learns a late-fusion
validation predictor.  In other words, it asks three related questions:
what is stable in each view, what is stable when all features compete, and how
much does each view help prediction on held-out tumours?

```{r integrated-fit}
lambda_list <- list(
  mrna  = lambda_mrna,
  mirna = lambda_mirna
)

multi_fit <- stabl_multiomic_train_validate(
  x_train_list    = list(mrna = X_train_mrna,  mirna = X_train_mirna),
  x_valid_list    = list(mrna = X_test_mrna,   mirna = X_test_mirna),
  y_train         = y_train,
  y_valid         = y_test,
  lambda_grid     = lambda_list,
  family          = "binomial",
  n_bootstraps    = 30L,
  artificial_type = "random_permutation",
  random_state    = 42L,
  early_fusion    = TRUE,
  late_fusion     = TRUE,
  n_iter_lf       = 500L
)
multi_fit
```

```{r integrated-results}
cat("mRNA (integrated):\n")
print(get_feature_names_out(multi_fit$fits$mrna))

cat("\nmiRNA (integrated):\n")
print(get_feature_names_out(multi_fit$fits$mirna))

cat("\nEarly fusion selected feature names:\n")
print(get_feature_names_out(multi_fit$early_fusion$fit))

cat("\nLate fusion omic weights:\n")
print(multi_fit$late_fusion$weights)
```

```{r integrated-paths, fig.alt = "Integrated mRNA and miRNA stability paths"}
print(plot_stabl_path(multi_fit$fits$mrna,  title = "Integrated - mRNA"))
print(plot_stabl_path(multi_fit$fits$mirna, title = "Integrated - miRNA"))
```

## 4. Validation performance

Feature stability is only one axis of evidence.  The held-out test set asks
whether the selected feature names and fusion weights carry predictive information
outside the training data.

```{r validation-performance}
lf_preds <- multi_fit$late_fusion$valid_predictions

plot_roc(y_true = y_test, y_preds = lf_preds,
         title = "Late fusion - ROC curve (test set)")
plot_prc(y_true = y_test, y_preds = lf_preds,
         title = "Late fusion - PRC curve (test set)")

pred_class <- as.integer(lf_preds >= 0.5)
conf_mat <- table(
  Predicted = factor(pred_class, levels = c(0, 1)),
  Actual    = factor(y_test, levels = c(0, 1))
)
print(conf_mat)

sensitivity <- conf_mat["1", "1"] / sum(conf_mat[, "1"])
specificity <- conf_mat["0", "0"] / sum(conf_mat[, "0"])
ber <- 1 - (sensitivity + specificity) / 2

cat(sprintf("\nSensitivity (Basal):     %.3f\n", sensitivity))
cat(sprintf("Specificity (non-Basal): %.3f\n", specificity))
cat(sprintf("Balanced Error Rate:     %.3f\n", ber))
```

```{r feature-boxplots, fig.alt = "Distribution of selected mRNA features"}
mrna_feats <- get_feature_names_out(multi_fit$fits$mrna)
if (length(mrna_feats) > 0) {
  boxplot_features(
    features = mrna_feats,
    x        = X_train_mrna,
    y        = y_train,
    title    = "Top mRNA features - training set",
    ncol     = min(3L, length(mrna_feats))
  )
}
```

## 5. Export results

For a vignette, exports go to `tempdir()` so rendering does not leave generated
files in the package source tree.  In a real analysis, these CSVs are the
bridge to notebooks, review tables, and downstream biological interpretation.

```{r export-results}
tcga_out_dir <- file.path(tempdir(), "stablr_tcga_results")
export_stabl_to_csv(multi_fit$fits$mrna,  path = file.path(tcga_out_dir, "mrna"))
export_stabl_to_csv(multi_fit$fits$mirna, path = file.path(tcga_out_dir, "mirna"))
list.files(tcga_out_dir, recursive = TRUE)
```

## 6. Comparison with mixOmics (sgccaDA and DIABLO)

The mixOmics ecosystem and `stablr` both address multi-omic classification, but
they optimise **different objectives**.  Treating them as interchangeable
classifiers misses the point: one searches for sparse latent factors that
separate classes; the other searches for **named features whose selection
frequency survives resampling and passes an FDP-aware threshold**.

### Scientific question

| Question | mixOmics (sgccaDA / DIABLO) | stablr |
|----------|----------------------------|--------|
| What is optimised? | Sparse loadings on latent components that maximise block-wise covariance with the outcome | Resampling stability of individual feature coefficients |
| What is the deliverable? | Components, block loadings, `selectVar()` lists tied to components | Explicit feature names, stability scores, FDP+ threshold |
| Integration mechanism | Joint multiblock PLS / sGCCA with a design matrix coupling blocks | Per-omic STABL, early/late fusion, or cooperative fusion (`rho`) |
| FDR / stability calibration | Sparsity via `keepX` tuning; no bootstrap stability or knockoff FDP+ | Knockoffs or random artificial features; FDP+ threshold sweep |
| Prediction | Native (`predict`, `centroids.dist`) | Requires a downstream refit on selected feature names (e.g. multinomial glmnet) |

### Objective strengths and limitations

**Choose mixOmics DIABLO when:**

- The goal is **integrative discrimination** with interpretable latent structure
  across correlated blocks.
- You want a mature tuning workflow (`tune.block.splsda`, design-matrix priors)
  and rich visualisation (`plotIndiv`, `plotVar`, circos plots).
- Classification (or block-supervised reduction) is the primary endpoint and
  component loadings are an acceptable feature representation.

**Choose stablr when:**

- The goal is a **reproducible sparse biomarker list** with an FDP+ diagnostic
  under the artificial-feature assumptions. The fitted threshold minimises the
  diagnostic curve; it is not a universal false-discovery guarantee.
- You need gaussian, binomial, multinomial, or Cox STABL on the same API, with
  grouped bootstrap for repeated measures.
- You want to compare integration strategies (per-omic, early, late, cooperative)
  under one stability-selection framework.
- You prefer staying in the glmnet / resampling idiom rather than PLS components.

**Neither replaces the other.**  On TCGA, DIABLO may win on outer-fold BER when
blocks share strong latent structure; stablr may surface features with higher
cross-fold recurrence that are not top DIABLO loadings.  Quantitative
head-to-head results live in `vignette("stablr-tcga-nestedcv")`; the full
benchmark protocol is documented in
`inst/analysis/BENCHMARK_PLAN.md`.

| Aspect | sgccaDA (Ch. 6) | DIABLO | stablr |
|--------|-----------------|--------|--------|
| Method | Sparse CCA + LDA | `block.splsda` + design matrix | Bootstrap-stability glmnet |
| Feature selection | L1 on components | `keepX` per block | FDP+ stability threshold |
| Integration | Joint latent space | Correlated latent components | Early, late, or cooperative fusion |
| Primary readout | BER + loadings | BER + `selectVar` | Recurrence + stability scores |
| Survival / Cox | Other mixOmics modules | Classification focus | Native `family = "cox"` |

Both sgccaDA and DIABLO target compact mRNA and miRNA signatures for subtype
classification.  STABL's distinctive contribution is FDP-aware stability
selection without assuming that omics must share a single latent space.

```{r session-info}
sessionInfo()
```
