---
title: "Cooperative Fusion for Multi-Omic Biomarker Discovery"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cooperative Fusion for Multi-Omic Biomarker Discovery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4
)
set.seed(42)
```

## Overview

Early fusion asks all features to compete in one wide matrix.  Late fusion lets
each omic make its own prediction and combines the answers afterward.
Cooperative fusion asks a subtler question: can the views be fitted together
without erasing their identities?

In `stablr`, cooperative fusion is built in natively (vendored from the
[multiview](https://cran.r-project.org/package=multiview) cooperative-learning
engine). Each view keeps its own coefficient path, but the paths are encouraged
to agree through a cooperation strength `rho`. The omics are neither fully
pooled nor fully isolated. Native v1 supports **gaussian** and **binomial**
families.

**When should you use cooperative fusion?**

| Strategy | How it works | Best when |
|---|---|---|
| **Per-omic** | One STABL fit per omic, independently. | Baseline; or when views are likely orthogonal. |
| **Early fusion** | Concatenate all omics, one joint STABL fit. | Omics are tightly correlated; you want a single feature list. |
| **Late fusion** | Per-omic fits, then learn a weighted blend of predictions. | Each omic may predict independently; you want omic weights. |
| **Cooperative fusion** | Joint multiview fit with cooperation strength `rho`; each view has its own coefficient path. | Omics share some signal structure; you want cross-view feature selection to be coherent. |

The cooperation strength `rho` controls how much the views are encouraged to
share signal.  At `rho = 0`, cooperative fusion reduces to independent fits. As
`rho` grows, the views are penalised toward a common prediction.  `stablr`
tunes `rho` over a user-supplied grid.

## Learning objectives

By the end, you should be able to:

1. Explain what `rho` controls in cooperative fusion and how to choose a grid.
2. Run `stabl_multiomic_train_validate()` with `cooperative_fusion = TRUE` and
   navigate the `$cooperative_fusion` result element.
3. Read the `print()` output for cooperative fusion details.
4. Understand which `cooperation_selection` and `cooperation_selector`
   combinations are supported and why.
5. Compare cooperative fusion feature selections to early and late fusion on the
   same dataset.
6. Use `stabl_multiomic_cv()` to evaluate cooperative fusion across outer folds.

---

## 1. Load the data

We use the same Onset of Labor (OOL) dataset as `vignette("stablr-multiomic")`:
CyTOF immune profiling and Proteomics measured on the same women, with
days-to-onset-of-labor (DOS) as the continuous outcome.  The examples use
compact grids so the vignette remains buildable; enlarge the bootstrap count
and `rho` grid for a final analysis.

```{r load}
library(stablr)
```

```{r load-data}
ool_train <- load_ool_data(split = "train")
ool_valid <- load_ool_data(split = "valid")

# Shape of each omic layer
lapply(ool_train$x_list, dim)
```

---

## 2. Build lambda grids

Cooperative fusion still needs a lambda grid for each omic.  We build one grid
per view with `auto_lambda_grid()` and pass the grids as a named list.  Twelve
lambda values are enough to demonstrate the workflow without making each
cooperative fit too expensive.

```{r lambda-grids}
lambda_list <- list(
  cytof      = auto_lambda_grid(
    ool_train$x_list$cytof, ool_train$y,
    family = "gaussian", n_lambda = 12
  ),
  proteomics = auto_lambda_grid(
    ool_train$x_list$proteomics, ool_train$y,
    family = "gaussian", n_lambda = 12
  )
)

# Number of lambda values per omic
vapply(lambda_list, nrow, integer(1L))
```

---

## 3. Running cooperative fusion

Activate cooperative fusion with `cooperative_fusion = TRUE` and a `rho` grid.
`stablr` fits one multiview model per `rho` value using shared inner CV folds,
then picks the best `rho` by the chosen `cooperation_type_measure`. This is
**sequential** tuning over `rho` (one full CV path per value), not a single
joint CV over `(rho, lambda)` pairs.

Key arguments:

| Argument | Default | Purpose |
|---|---|---|
| `rho` | `0` | Scalar or vector of cooperation strengths to try. |
| `cooperation_selection` | `"cv"` | `"cv"` (inner CV) or `"validation"` (held-out set). |
| `cooperation_selector` | `"lambda.min"` | Which lambda to pick: `"lambda.min"` or `"lambda.1se"` (CV only). |
| `cooperation_type_measure` | `"default"` | Tuning metric; defaults to `"mse"` for Gaussian. |
| `cooperation_nfolds` | `5L` | Number of inner CV folds when `cooperation_selection = "cv"`. |

```{r cooperative-fit}
cf_fit <- stabl_multiomic_train_validate(
  x_train_list          = ool_train$x_list,
  y_train               = ool_train$y,
  lambda_grid           = lambda_list,
  x_valid_list          = ool_valid$x_list,
  y_valid               = ool_valid$y,
  family                = "gaussian",
  n_bootstraps          = 30L,
  artificial_type       = "random_permutation",
  random_state          = 42L,
  cooperative_fusion    = TRUE,
  rho                   = c(0, 0.25, 0.5),
  cooperation_selection = "cv",
  cooperation_selector  = "lambda.min",
  cooperation_nfolds    = 3L
)
```

`print()` shows every active branch.  When cooperative fusion is present, a
`Cooperative fusion:` block appears at the bottom:

```{r print-fit}
cf_fit
```

**Reading the cooperative fusion block:**

- **`selection`** --- how `rho` was tuned: `"cv"` here.
- **`rho (chosen)`** --- the `rho` value selected by cross-validation.  A value
  of `0` means the specified metric and folds preferred independent fits within
  the supplied grid.  A larger value means that the same tuning surface preferred
  some coupling between views; it is not by itself evidence of biological
  cooperation.
- **`selector`** --- which lambda rule was applied (`lambda.min`).
- **`type.measure`** --- the CV metric used for tuning (e.g., `mse`).
- **`score`** --- the CV score at the chosen `rho` and `lambda`.
- **`selected feats`** --- total number of features selected across all views.

---

## 4. Navigating `$cooperative_fusion`

The `cooperative_fusion` element contains the selected model and the tuning
surface that led to it.

```{r cf-element}
cf <- cf_fit$cooperative_fusion

# Chosen hyperparameters
cat("Chosen rho:     ", cf$rho, "\n")
cat("Chosen lambda:  ", round(cf$selected_lambda, 4), "\n")
cat("Selector:       ", cf$selector, "\n")
cat("CV score (MSE): ", round(cf$score, 4), "\n")
```

### Selected features per view

`$selected_features` is a named list with one character vector per omic:

```{r cf-selected}
cat("CyTOF features selected by cooperative fusion:\n")
print(cf$selected_features$cytof)

cat("\nProteomics features selected by cooperative fusion:\n")
print(cf$selected_features$proteomics)
```

### Predictions

`$train_predictions` and `$valid_predictions` hold the multiview predictions
from the chosen model:

```{r cf-predictions}
cat("First 6 training predictions:\n")
print(head(cf$train_predictions))

cat("\nFirst 6 validation predictions:\n")
print(head(cf$valid_predictions))
```

### Tuning diagnostics

`$diagnostics` records the CV score at every `rho` in the grid, so the tuning
choice can be inspected rather than treated as a black box:

```{r cf-diagnostics}
# One row per rho value
print(cf$diagnostics[, c("rho", "lambda", "metric_value", "selected")])
```

The `selected` column marks the winning `rho`.  If the surface is flat, with
similar `metric_value` across `rho`, the omics are not gaining much from
cooperation and `rho = 0` is a defensible choice.

---

## 5. Policy constraints

Some combinations are rejected on purpose.  These guards prevent analyses that
look reasonable in code but do not have a valid tuning interpretation.

### Constraint 1 - `lambda.1se` requires CV mode

`lambda.1se` applies the "one-standard-error" rule and is only well-defined
when cross-validation is used to estimate the error surface.  Requesting it
with `cooperation_selection = "validation"` raises an error:

```{r guard-lambda1se, error = TRUE}
stabl_multiomic_train_validate(
  x_train_list          = ool_train$x_list,
  y_train               = ool_train$y,
  lambda_grid           = lambda_list,
  x_valid_list          = ool_valid$x_list,
  y_valid               = ool_valid$y,
  family                = "gaussian",
  n_bootstraps          = 3L,
  artificial_type       = NULL,
  cooperative_fusion    = TRUE,
  rho                   = 0,
  cooperation_selection = "validation",
  cooperation_selector  = "lambda.1se"
)
```

### Constraint 2 - Cox and Poisson families are not supported in native v1

Native cooperative fusion in `stablr` currently supports **gaussian** and
**binomial** families only. Poisson and Cox cooperative paths are deferred to a
future release:

```{r guard-cox-family, error = TRUE}
library(survival)

set.seed(42)
n <- 20L
ids <- paste0("s", seq_len(n))
x_a <- matrix(rnorm(n * 4), nrow = n,
               dimnames = list(ids, paste0("a", seq_len(4))))
x_b <- matrix(rnorm(n * 4), nrow = n,
               dimnames = list(ids, paste0("b", seq_len(4))))
y_surv <- Surv(time = rexp(n), event = rbinom(n, 1L, 0.7))
rownames(y_surv) <- ids

stabl_multiomic_train_validate(
  x_train_list          = list(omic_a = x_a, omic_b = x_b),
  y_train               = y_surv,
  lambda_grid           = data.frame(lambda = c(0.2, 0.1)),
  family                = "cox",
  n_bootstraps          = 2L,
  artificial_type       = NULL,
  cooperative_fusion    = TRUE,
  rho                   = 0
)
```

---

## 6. Built-in cooperative engine

Cooperative fusion is built into `stablr`; no separate `multiview` install is
required. The engine is vendored from CRAN `multiview` v1.0 (GPL-2; see
`inst/COPYING.cooperative`) and is enabled with `cooperative_fusion = TRUE`.

Setting `cooperative_fusion = FALSE`, the default, runs only the per-omic STABL
paths and never invokes the cooperative engine.

---

## 7. Comparing cooperative, early, and late fusion

The most useful comparison is often side by side.  Run all three strategies in
one call, then compare the selected features:

```{r all-strategies}
all_fit <- stabl_multiomic_train_validate(
  x_train_list          = ool_train$x_list,
  y_train               = ool_train$y,
  lambda_grid           = lambda_list,
  x_valid_list          = ool_valid$x_list,
  y_valid               = ool_valid$y,
  family                = "gaussian",
  n_bootstraps          = 30L,
  artificial_type       = "random_permutation",
  random_state          = 42L,
  early_fusion          = TRUE,
  late_fusion           = TRUE,
  n_iter_lf             = 250L,
  cooperative_fusion    = TRUE,
  rho                   = c(0, 0.25, 0.5),
  cooperation_selection = "cv",
  cooperation_selector  = "lambda.min",
  cooperation_nfolds    = 3L
)
```

```{r compare-print}
# print() now shows all four branches
all_fit
```

Compare feature sets across strategies:

```{r compare-features}
ef_features  <- all_fit$early_fusion$selected_features
cf_features  <- unique(unlist(all_fit$cooperative_fusion$selected_features,
                              use.names = FALSE))
per_omic_features <- unique(unlist(all_fit$selected_features,
                                   use.names = FALSE))

cat("Early fusion selected:      ", length(ef_features), "features\n")
cat("Cooperative fusion selected:", length(cf_features), "features\n")
cat("Per-omic union selected:    ", length(per_omic_features), "features\n")

cat("\nIn cooperative but not early fusion:\n")
print(setdiff(cf_features, ef_features))

cat("\nIn early fusion but not cooperative:\n")
print(setdiff(ef_features, cf_features))
```

**Interpreting differences:** early fusion sees all features simultaneously in
one penalised model; cooperative fusion fits per-view paths that are encouraged
to agree.  Features unique to cooperative fusion may carry cross-view signal
that is diluted in the concatenated early-fusion space.

### Relation to mixOmics DIABLO

Cooperative fusion is the closest `stablr` analogue to multiblock integration,
but it is **not** equivalent to DIABLO:

| | Cooperative fusion (`rho`) | DIABLO |
|---|---------------------------|--------|
| Optimisation | Penalised regression paths per view, coupled by cooperation penalty | Sparse PLS components with block design matrix |
| Output | Per-view coefficient paths and selected features | Latent components + block loadings |
| Tuning | `rho`, `lambda`, `cooperation_nfolds` | `ncomp`, `keepX`, design matrix entries |
| Families (v1) | gaussian, binomial | Classification (sPLS-DA) |
| Stability / FDP | STABL bootstrap + knockoff layer | Component sparsity only |

Use cooperative fusion when you want **stable sparse features per omic** with
cross-view agreement.  Use DIABLO when you want **joint latent factors** for
classification and mixOmics visualisation.  Head-to-head TCGA benchmarks that
include early fusion (and future cooperative arms) are described in
`inst/analysis/BENCHMARK_PLAN.md`.

---

## 8. Cross-validation with cooperative fusion

Use `stabl_multiomic_cv()` when no predefined validation split is available.
Cooperative fusion parameters propagate to every outer fold.  This is a
realistic analysis pattern, but it multiplies cost by the number of outer
folds, so the code is shown but not evaluated during vignette builds:

```{r cv-fit, eval = FALSE}
# Long-running vignette branch; exercised by test-multiomic-workflows.R.
cv_fit <- stabl_multiomic_cv(
  x_list                = ool_train$x_list,
  y                     = ool_train$y,
  lambda_grid           = lambda_list,
  v                     = 3L,
  family                = "gaussian",
  n_bootstraps          = 20L,
  artificial_type       = "random_permutation",
  random_state          = 42L,
  cooperative_fusion    = TRUE,
  rho                   = c(0, 0.3),
  cooperation_selection = "cv",
  cooperation_selector  = "lambda.min",
  cooperation_nfolds    = 3L
)

cv_fit
```

The `$diagnostics` data frame includes cooperative-specific columns
(`cooperative_rho`, `cooperative_lambda`, `cooperative_score`, etc.), making it
possible to inspect how `rho` was selected in each outer fold:

```{r cv-diagnostics, eval = FALSE}
# Output contract exercised by test-multiomic-workflows.R.
print(cv_fit$diagnostics[, c("fold", "omic", "n_selected",
                              "cooperative_rho", "cooperative_score")])
```

---

## 9. Choosing `rho` and design guidance

The useful `rho` grid is usually modest.  Treat it as a scientific sensitivity
analysis: did the views need to cooperate, and if so, how strongly?

| Decision | Guidance |
|---|---|
| **Grid density** | Start with `c(0, 0.1, 0.3, 0.5)`.  Add finer points only if the CV surface has a clear non-zero minimum. |
| **Grid upper bound** | Values > 0.5 can over-penalise disagreement between views; rarely beneficial unless views are near-identical. |
| **`lambda.min` vs `lambda.1se`** | `lambda.1se` gives a sparser model at the cost of a higher CV error; prefer `lambda.min` when prediction is the goal. |
| **`cooperation_nfolds`** | 5 (default) is suitable for n >= 50.  Use 3 for smaller datasets.  Must be >= 3. |
| **`cooperation_selection = "validation"`** | Use when you have a true held-out set and want to tune `rho` and `lambda` together on it. |
| **Native v1 families** | Cooperative fusion supports `family = "gaussian"` and `"binomial"` only. |
| **Binomial tuning metrics** | Try `cooperation_type_measure = "auc"` or `"class"` when the outcome is binary. |
| **rho = 0 wins** | If the chosen `rho` is always 0, the chosen metric prefers independent fits under this grid and fold design.  Treat that as a prompt for sensitivity analysis, not proof that the omics share no useful structure. |

---

## Next steps

- See `vignette("stablr-multiomic")` for per-omic, early, and late fusion on
  the same OOL dataset.
- See `vignette("stablr-intro")` for single-omic STABL and learner options.
- Pass `groups` to `stabl_multiomic_train_validate()` to prevent leakage in
  longitudinal designs; grouped bootstrapping and outer-fold construction both
  respect this argument.
- For binomial cooperative fusion, set `family = "binomial"` and consider
  `cooperation_type_measure = "auc"` or `"class"`.
- See `vignette("stablr-advanced")` for Cox survival STABL, knockoffs,
  grouped bootstrap, reproducibility metrics, export helpers, and outer CV.
- For mixOmics DIABLO comparison on TCGA, see `vignette("stablr-tcga-nestedcv")`.
