---
title: "Introduction to stablr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to stablr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4
)
set.seed(42)
```

## What is STABL?

A high-dimensional biomarker screen begins with an uncomfortable fact: there
are many more tempting variables than there are samples.  A sparse model can
still choose a short list, but a single fit does not tell us whether the list is
sturdy or merely lucky.

STABL (Stability-based Thresholding and Aggregation for Biomarker discovery via
LASSO) turns that one-shot question into a repeated experiment.  It resamples
the data, fits a sparse learner across a penalty grid, records how often each
feature appears, and calibrates the result against artificial decoy features.
The central object is therefore not one coefficient vector.  It is a frequency:
how often does this feature survive when the data are perturbed?

The `stablr` package is R-native without a Python runtime and built around the `glmnet`
ecosystem.  This introductory vignette uses small simulated datasets,
random-permutation decoys, and modest bootstrap counts so the examples remain
quick enough for routine package builds.

The workflow is deliberately simple:

1. Prepare a numeric feature matrix `x` and an outcome vector `y`.
2. Build a lambda grid with `auto_lambda_grid()`.
3. Fit STABL with `stabl_fit()`.
4. Inspect selected features, stability scores, and diagnostic plots.

The fitted object is a ranked feature-selection result rather than a final
prediction model.  It records which features cross the stability threshold, how
often each feature was selected, and how those selections compare with the
decoys.

## Installation

```{r install, eval = FALSE}
# Interactive installation protocol; not a package API example.
# From the repository root after cloning:
devtools::install(".")
```

```{r load}
library(stablr)
```

## Scope

This vignette is the small door into the package.  It covers the single-omic
case:

1. preparing the input objects expected by `stabl_fit()`;
2. building a compact penalty grid with `auto_lambda_grid()`;
3. fitting a small binary or regression example;
4. extracting selected feature names (`get_feature_names_out()`), logical
   support masks (`get_support()`), and stability scores (`get_importances()`).
5. reading the stability-path and FDP diagnostic plots;
6. switching between lasso, adaptive lasso, and elastic-net base learners.

## Basic input shape

The method expects ordinary R objects.  The important convention is not exotic:
samples are rows, features are columns, and the outcome is aligned to the same
sample order.

| Object | Expected shape | Notes |
|---|---|---|
| `x` | numeric matrix or matrix-like object, samples in rows and features in columns | Column names become feature names in output. |
| `y` | outcome vector with one value per row of `x` | Use `0`/`1` or a factor-like outcome for `family = "binomial"`; use numeric values for `family = "gaussian"`. |
| `lambda_grid` | data frame returned by `auto_lambda_grid()` | Each row defines a sparse model setting evaluated during bootstrapping. |

If `x` has row names and `y` has names, both must contain the same unique sample
IDs. Their order may differ: `stablr` aligns the outcome to predictor rows by
ID. In biomarker work, a silent sample mismatch is worse than a model error
because it can look perfectly plausible.

---

## 1. Binary classification

### Simulated data

The first experiment is binary classification.  We build a small world in
which the truth is known: 120 samples, 24 candidate features, and only five
real signals (`B1` to `B5`).  The remaining features are noise.  This lets us
ask the cleanest possible question: does stability selection recover the
features that were planted?

```{r sim-binary}
n_bin <- 120; p_bin <- 24; n_signal_bin <- 5
signal_features_bin <- paste0("B", seq_len(n_signal_bin))

X_bin <- matrix(rnorm(n_bin * p_bin), nrow = n_bin,
                dimnames = list(paste0("s", seq_len(n_bin)),
                                paste0("B", seq_len(p_bin))))

# True signal in B1-B5; all other features are independent noise.
beta_bin       <- c(rep(1.8, n_signal_bin), rep(0, p_bin - n_signal_bin))
log_odds       <- X_bin %*% beta_bin
y_bin          <- rbinom(n_bin, size = 1, prob = plogis(log_odds))
names(y_bin)   <- rownames(X_bin)
table(y_bin)

signal_features_bin
```

The `table()` output is a quick check on class balance.  A badly imbalanced
screen can still be useful, but its selected features need more care in
interpretation.

### Fitting STABL

There are two moving parts.  First, `auto_lambda_grid()` asks `glmnet` for a
data-dependent penalty path.  Each row of the grid becomes one sparse model
configuration.  Second, `stabl_fit()` repeats the fit across bootstrap samples
and counts feature selections.

The main arguments are:

| Argument | What it does |
|---|---|
| `family` | Passed to `glmnet`; `"binomial"` for 0/1 outcome, `"gaussian"` for continuous. |
| `n_bootstraps` | Number of bootstrap resamples.  The examples use 40-50; final analyses usually need hundreds. |
| `artificial_type` | Decoy feature type used for FDP calibration. |
| `random_state` | Integer seed for full reproducibility. |

Supported artificial-feature settings are:

| Value | Meaning |
|---|---|
| `"random_permutation"` | Column-permuted decoys; no optional dependency. |
| `"knockoff"` | Fixed-X knockoffs via the optional `knockoff` package. |
| `"knockoff_equi"` | Model-X equicorrelated Gaussian knockoffs. |
| `"knockoff_mvr"` | Model-X MVR Gaussian knockoffs. |
| `NULL` | No artificial features; FDP diagnostics are unavailable. |

All knockoff modes fail closed above 3,000 predictors because the package does
not claim a globally valid chunked construction. Random-permutation controls
do not have that dimensional restriction.

Scalar tuning arguments such as `n_bootstraps` and `random_state` must be
finite integer-like scalars.  Plot titles, output paths, and file-format
arguments must be non-empty character scalars.

The example uses `n_bootstraps = 50` and keeps the strong-penalty end of a
20-point lambda path.  This keeps the runtime short and the selected set
compact.  Lower penalties admit more features, which can be useful for
diagnostics but can hide a small planted signal in a short introductory run.
Final analyses should use more bootstrap resamples and should check that the
selected features are not an artifact of one narrow grid choice.

```{r fit-binary}
lambda_bin_full <- auto_lambda_grid(
  X_bin, y_bin, family = "binomial", n_lambda = 20
)
lambda_bin <- lambda_bin_full[seq_len(6), , drop = FALSE]

fit_bin <- stabl_fit(
  x               = X_bin,
  y               = y_bin,
  lambda_grid     = lambda_bin,
  family          = "binomial",
  n_bootstraps    = 50L,
  artificial_type = "random_permutation",
  random_state    = 42L
)

fit_bin
```

**Reading the printed summary.**  The output reports three key numbers:

- **Stability threshold** --- the minimum selection frequency a feature must
  reach to be called a biomarker (e.g. `0.7` means selected in >= 70% of
  bootstraps).
- **n selected** --- features passing the threshold; for our 5-signal data you
  expect a small number close to 5.
- **FDP+ estimate** --- estimated upper bound on the false-discovery proportion
  among selected features.  With only five planted signals and the additive
  `+1` FDP+ correction, the best possible value for a perfect five-feature
  selection is `1 / 5 = 0.2`; larger real analyses can support smaller values.

In real data, these numbers are a screening summary, not a verdict.  The most
convincing fits have a selected set of plausible size, a visible gap between
selected and unselected stability scores, and features that survive domain
review.

### Inspecting results

`get_feature_names_out()` gives the selected biomarker names.  `get_support()`
returns the same decision as a named logical vector, which is useful when the
original matrix needs to be subset.  `get_importances()` keeps the fuller story:
a stability score for every feature, sorted from most to least persistent.

Because this is a simulation, we can compare the selected set with the planted
truth.  In a real experiment, the analogous evidence comes from independent
validation, mechanism, and the shape of the stability-score gap.

```{r results-binary}
# Features selected at the STABL threshold
selected_bin <- get_feature_names_out(fit_bin)
selected_bin

# Subset new data to the selected columns
X_bin_selected <- transform_stabl(fit_bin, X_bin)
dim(X_bin_selected)

# Compare the selected set against the known simulation truth
setdiff(signal_features_bin, selected_bin)  # missed planted features
setdiff(selected_bin, signal_features_bin)  # selected noise features

# Stability scores for all features (sorted descending)
head(get_importances(fit_bin), 10)
```

The two `setdiff()` calls are a luxury of simulation.  They show what STABL is
being asked to do: separate repeatable signal from features that only look good
once.

### Visualisation

**Stability path.**  Each curve is one feature's selection frequency along the
lambda path.  A useful fit usually has a small group of curves that lift away
from the background.  In this simulation, `B1` to `B5` should form that upper
band.

```{r plot-binary, fig.alt = "Stability path for binary classification"}
plot_stabl_path(fit_bin, title = "Binary classification - stability path")
```

The plotting helpers return ordinary `ggplot` objects, so the same figure can
be displayed, modified, or saved.

```{r save-path-plot, eval = FALSE}
# Optional file writing; plot contract exercised by test-phase7.R.
path_plot <- plot_stabl_path(fit_bin, title = "Binary classification - stability path")
ggplot2::ggsave("stability-path.png", path_plot, width = 7, height = 5)
```

**FDP+ diagnostic (`plot_fdr_graph()`).** The function retains its historical
name, but the plotted quantity is FDP+. At threshold $t$, the numerator is the
artificial-feature exceedance count adjusted for the injected proportion, plus
one; the denominator is the real-feature exceedance count, bounded below by
one. The fitted threshold is the first grid value attaining the minimum FDP+.
The vertical dashed line marks that threshold, while the horizontal line is a
visual target only. In this tiny five-signal simulation the FDP+ lower bound is
`1 / 5 = 0.2`, so the graph is a calibration diagnostic, not a universal
false-discovery guarantee.

```{r fdr-binary, fig.alt = "FDR estimate graph for binary classification"}
plot_fdr_graph(fit_bin, title = "Binary classification - FDR estimate")
```

```{r save-fdr-plot, eval = FALSE}
# Optional file writing; plot contract exercised by test-phase7.R.
fdr_plot <- plot_fdr_graph(fit_bin, title = "Binary classification - FDR estimate")
ggplot2::ggsave("fdr-graph.png", fdr_plot, width = 7, height = 5)
```

---

## 2. Regression

### Simulated data

Regression changes the question from class membership to a continuous outcome.
The structure stays the same: a known signal is planted, noise is added, and
STABL is asked to find the features that keep reappearing.  Here the true
features are `R1` to `R5`.

```{r sim-reg}
n_reg <- 160; p_reg <- 24; n_signal_reg <- 5
signal_features_reg <- paste0("R", seq_len(n_signal_reg))

X_reg <- matrix(rnorm(n_reg * p_reg), nrow = n_reg,
                dimnames = list(paste0("s", seq_len(n_reg)),
                                paste0("R", seq_len(p_reg))))
beta_reg     <- c(rep(3, n_signal_reg), rep(0, p_reg - n_signal_reg))
signal_reg   <- as.numeric(X_reg %*% beta_reg)
y_reg        <- signal_reg + rnorm(n_reg, sd = 0.5)
names(y_reg) <- rownames(X_reg)

signal_features_reg
```

### Fitting STABL

The code mirrors the binary example.  Moving from classification to regression
requires changing `family`, rebuilding the lambda grid for the new outcome, and
refitting.

```{r fit-reg}
lambda_reg_full <- auto_lambda_grid(
  X_reg, y_reg, family = "gaussian", n_lambda = 20
)
lambda_reg <- lambda_reg_full[seq_len(6), , drop = FALSE]

fit_reg <- stabl_fit(
  x               = X_reg,
  y               = y_reg,
  lambda_grid     = lambda_reg,
  family          = "gaussian",
  n_bootstraps    = 50L,
  artificial_type = "random_permutation",
  random_state    = 42L
)

fit_reg
```

### Results

The strongest scores should concentrate on `R1` to `R5`.  Their exact scores
need not match the binary example because the Gaussian loss changes the
effective sparsity of each bootstrap model.

```{r results-reg}
selected_reg <- get_feature_names_out(fit_reg)
selected_reg

setdiff(signal_features_reg, selected_reg)  # missed planted features
setdiff(selected_reg, signal_features_reg)  # selected noise features

head(get_importances(fit_reg), 10)
```

The selected feature names are the thresholded screening result.  The full
importance vector is often just as useful, because features just below the
threshold may suggest a continuum rather than a sharp boundary.

### Visualisation

The stability path should echo the binary case: the planted features lift away
from the background.  Differences in curve height reflect the model family, not
a change in the definition of stability.

```{r plot-reg, fig.alt = "Stability path for regression"}
plot_stabl_path(fit_reg, title = "Regression - stability path")
```

---

## 3. Learner variants

The default base learner is lasso.  Other sparse learners are useful when the
geometry of the feature space changes.  The examples below reuse the binary
simulation so the learner choice is the only moving part.

### Adaptive lasso

The adaptive lasso re-weights the L1 penalty.  Features with small coefficients
in a preliminary ridge fit are penalised more heavily, which can sharpen the
contrast between persistent signal and unstable noise.

**When to use it:** prefer adaptive lasso when you expect a sparse true signal
and want a tighter biomarker list with fewer borderline features.

**What to expect:** B1-B5 should remain among the strongest-ranked features,
and `get_importances()` will typically show higher scores for true features
compared to plain lasso.  Borderline noise features can still appear because
this variant chunk uses a broader introductory lambda grid.

```{r adaptive-lasso}
lambda_ada <- auto_lambda_grid(X_bin, y_bin, family = "binomial", n_lambda = 10)

fit_ada <- stabl_fit(
  x               = X_bin,
  y               = y_bin,
  lambda_grid     = lambda_ada,
  base_learner    = "adaptive_lasso",
  family          = "binomial",
  n_bootstraps    = 40L,
  artificial_type = "random_permutation",
  random_state    = 42L
)

get_feature_names_out(fit_ada)
sum(get_support(fit_ada))
```

The count from `sum(get_support(fit_ada))` shows how compact the selected set
has become.  `get_importances(fit_ada)` keeps the full ranking.

### Elastic net

Elastic net interpolates between lasso (`alpha = 1`) and ridge (`alpha = 0`).
The L2 component lets correlated features travel together, which is useful when
biomarkers form modules rather than isolated hits.

Passing multiple `l1_ratio` values to `auto_lambda_grid()` makes STABL sweep
the (lambda, alpha) grid jointly, exploring both the sparsity level and the
L1/L2 mix in every bootstrap.

**When to use it:** prefer elastic net over plain lasso when features are highly
correlated and you want to avoid arbitrarily picking one representative from a
correlated group.

```{r elastic-net}
lambda_en <- auto_lambda_grid(
  X_bin, y_bin,
  family = "binomial", n_lambda = 8,
  l1_ratio = c(0.5, 0.9)
)

fit_en <- stabl_fit(
  x               = X_bin,
  y               = y_bin,
  lambda_grid     = lambda_en,
  base_learner    = "elastic_net",
  family          = "binomial",
  n_bootstraps    = 40L,
  artificial_type = "random_permutation",
  random_state    = 42L
)

get_feature_names_out(fit_en)
```

---

## Next steps

This vignette stays close to the core idea: repeat the sparse fit, count what
survives, and calibrate that survival against decoys.  From here:

- Use more bootstraps for final analyses, typically hundreds rather than the
  small values used here.
- Use `vignette("stablr-multiomic")` when your data contain multiple omics
  views.
- Use `?save_stabl_results` when you are ready to export scores, plots, and
  selected-feature tables to disk.
